---
title: "Cyclistic Divvy Bikes Case Study"
author: "Sean Mattison"
date: "2023-10-13"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Cyclistic_Divvy_Trips_2019_Analysis ###

### Using the Divvy_Trips_2019 dataset for the case study. The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct
simple analysis to help answer the key question: “In what ways do members and casual riders use Divvy bikes differently?” The data is publicly available and has been provided by Motivate International Inc. under an appropriate license agreement.

# Check and set working directory to simplify calls to data
```{r Check and set working directory to simplify calls to data}
getwd()
setwd("/Users/seanmattison1/Desktop/divvy_bike_data")

```

# Install required packages
```{r Install required packages}
options(repos = structure(c(CRAN = "https://cran.r-project.org")))
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
install.packages("tinytex")
library(tidyverse) # helps wrangle data
library(lubridate) # helps wrangle date attributes
library(ggplot2) # helps visualize data
library(dplyr) # manipulation and transformation
library(tinytex)
```
#======================
#STEP 1: COLLECT DATA
#======================
# Collect data, upload Divvy dataset
```{r Upload .csv files}
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
```
#====================================================
#STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
#====================================================
# Compare column names in each of the files
```{r Compare column names}
colnames(q1_2019)
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
```
# Rename columns to make them consistent with q1_2019, q2_2019, and q3_2019
```{r Clean up column names}
(q2_2019 <- rename(q2_2019
			,trip_id = "01 - Rental Details Rental ID"
			,start_time = "01 - Rental Details Local Start Time"
			,end_time = "01 - Rental Details Local End Time"
			,bikeid = "01 - Rental Details Bike ID"
			,tripduration = "01 - Rental Details Duration In Seconds Uncapped"
			,from_station_id = "03 - Rental Start Station ID"
			,from_station_name = "03 - Rental Start Station Name"
			,to_station_id = "02 - Rental End Station ID"
			,to_station_name = "02 - Rental End Station Name"
			,usertype = "User Type"
			,gender = "Member Gender"
			,birthyear = "05 - Member Details Member Birthday Year"))


```
# Inspect the data frames and look for incongruencies
```{r check out the struckture of each data set}
str(q1_2019)
str(q2_2019)
str(q3_2019)
str(q4_2019)
```
# Stack indavidual quarter's data frames into one big data frame
```{r Aggregate data}
all_trips <- bind_rows(q1_2019,q2_2019,q3_2019,q4_2019)
```
# Remove birthyear and gender fields as this data is not relevant to our analysis
```{r Remove irrelevant data}
all_trips <- all_trips %>%
	select(-c(gender, birthyear, tripduration))
```
#=====================================================
#STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
#=====================================================
# Inspect the new table that has been created
```{r Inspect new table}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
tail(all_trips)
str(all_trips)
summary(all_trips)
```

# Remove rows where the 'end_time' or 'start_time' column contains the year 2020
```{r Remove inconsistent data}
all_trips <- all_trips %>%
   filter(year(start_time) != 2020 & year(end_time) != 2020)
```
# In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"
# Before 2020, Divvy used different labels for these two types of riders...we will want to make our data frame consistent with their currennt nomenclature, including all column names aswell

# Reanme all column names for better understanding and consistency with future data sets
```{r}
all_trips <- rename(all_trips
			,ride_id=trip_id
			,started_at=start_time
			,ended_at=end_time
			,rideable_type=bikeid
			,start_station_id=from_station_id
			,start_station_name=from_station_name
			,end_station_id=to_station_id
			,end_station_name=to_station_name
			,member_casual=usertype)

```
# Begin by seeing how many observations fall under each usertype
```{r Changing variable names for data consistency}
table(all_trips$member_casual)


```

# In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"
```{r Mutating variables}
all_trips <- all_trips %>%
	mutate(member_casual = recode(member_casual
		,"Subscriber" = "member"
		,"Customer" = "casual"))


```

# Check to make sure the proper number of observations were reassigned
```{r Inspect data to ensure numbers match}
table(all_trips$member_casual)

```

# Add columns that list the date, month, day, and year of each ride, this will allow aggregated ride data for each month, day, or year...
# The default format is yyyy-mm-dd
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date),"%m")
all_trips$day <- format(as.Date(all_trips$date),"%d")
all_trips$year <- format(as.Date(all_trips$date),"%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date),"%A")

```

# Add a "ride_length" calculation to all_trips (in seconds)
```{r Add ride length column}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at) 

```
# Inspect the structure of the columns
```{r}
str(all_trips)
```
# Convert "ride_length" from Factor to numeric so calculations can be performed on the data
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```
# Remove "bad" data
# The data frame includes a few hundred entries when bikes were taken out of docks nd checked for quality by Divvy or ride_length was negative.
# Since data is being removed, a new version of the data frame (v2) will be created
```{r}
all_trips <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
# Calculate IQR and define upper bound for outliers
```{r}
Q1 <- quantile(all_trips$ride_length, 0.25)
Q3 <- quantile(all_trips$ride_length, 0.75)
IQR <- Q3 - Q1
upper_bound <- Q3 + 1.5 * IQR

```
# Identify outliers
```{r}
outliers <- all_trips$ride_length > upper_bound
```
# Remove outliers
```{r}
all_clean <- all_trips[!outliers, ]


```
#======================================
#STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
#======================================
# Descriptive analysis on ride_length (all figures in seconds)
```{r}
mean(all_clean$ride_length)	
median(all_clean$ride_length)
max(all_clean$ride_length)
min(all_clean$ride_length)
summary(all_clean$ride_length)

```
# Compare members and casual users
```{r}
aggregate(all_clean$ride_length~all_clean$member_casual, FUN=mean)
aggregate(all_clean$ride_length~all_clean$member_casual, FUN=median)
aggregate(all_clean$ride_length~all_clean$member_casual, FUN=max)
aggregate(all_clean$ride_length~all_clean$member_casual, FUN=min)


```
# See the average ride time by each day for members vs casual users
```{r}
aggregate(all_clean$ride_length~all_clean$member_casual+all_clean$day_of_week, FUN=mean)
```
# Fix the order of the days of the week
```{r}
all_clean$day_of_week <- ordered(all_clean$day_of_week, levels=c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))
```
# Analyze ridership data by type and weekday
```{r}
all_clean %>%
	mutate(weekday=wday(started_at, label=TRUE)) %>% # creates weekday field using wday()
	group_by(member_casual, weekday) %>% # groups by usertype and weekday
	summarise(number_of_rides=n() # calculates the number of ridesand average duration
	,average_duration=mean(ride_length)) %>% # calculates the average duration
	arrange(member_casual, weekday) # sorts

```
# Visualize the number of rides by rider type
```{r}
all_clean %>%
	mutate(weekday=wday(started_at, label=TRUE)) %>%
	group_by(member_casual, weekday) %>%
	summarise(number_of_rides=n()
		,average_duration=mean(ride_length)) %>%
	arrange(member_casual, weekday) %>%
	ggplot(aes(x=weekday,y=number_of_rides,fill=member_casual))+
	geom_col(position="dodge")

```

# Visualization for average duration
```{r}
all_clean %>%
	mutate(weekday=wday(started_at, label=TRUE)) %>%
	group_by(member_casual, weekday) %>%
	summarise(number_of_rides=n()
		,average_duration=mean(ride_length)) %>%
	arrange(member_casual, weekday) %>%
	ggplot(aes(x=weekday,y=average_duration,fill=member_casual))+
	geom_col(position="dodge")	

```

#==================================================
#STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS
#==================================================
# Create a csv file that will be visualized in Tableau
```{r}
counts <- aggregate(all_clean$ride_length~all_clean$member_casual+
all_clean$day_of_week,FUN=mean)
write.csv(counts,file='~/Desktop/divvy_bike_data/avg_ride_length.csv')
write.csv(all_clean, file = '~/Desktop/divvy_bike_data/all_trips_clean.csv')
```
